\begin{proposition}[Disentanglement under Orthogonal Controllers]
\label{prop:orthogonal_fusion}
Let \(\tilde{s}^{(1)}, \tilde{s}^{(2)} \in \mathbb{R}^d\) be orthogonal controller vectors such that \(\langle \tilde{s}^{(1)}, \tilde{s}^{(2)} \rangle = 0\), and let the layer \(\ell\) hidden state be modulated by additive controller fusion:
\[
h^\ell = f(x^\ell) + W_m^\ell \tilde{s},
\]
where \(W_m^\ell \in \mathbb{R}^{d' \times d}\) is the injection weight matrix for fusion mode \(m\), and \(f(\cdot)\) is the controller-independent component. Assume the final output logits are computed via a linear decoder:
\[
\mathcal{F}_m(x, \tilde{s}) = V h^L,
\]
where \(V \in \mathbb{R}^{C \times d'}\) projects to logits over \(C\) classes. If \(W_m^\ell\) is full rank and \(V W_m^\ell\) has spectral norm bounded below by \(\sqrt{\epsilon} > 0\), then:
\[
\left\| \mathcal{F}_m(x, \tilde{s}^{(1)}) - \mathcal{F}_m(x, \tilde{s}^{(2)}) \right\|_2^2 \geq \epsilon \cdot \left\| \tilde{s}^{(1)} - \tilde{s}^{(2)} \right\|_2^2.
\]
\end{proposition}

\begin{proof}
\textbf{Step 1:} Express the difference in output logits.
We compute the difference in output logits:
\[
\Delta := \mathcal{F}_m(x, \tilde{s}^{(1)}) - \mathcal{F}_m(x, \tilde{s}^{(2)}) = V W_m^\ell (\tilde{s}^{(1)} - \tilde{s}^{(2)}).
\]

\textbf{Step 2:} Compute the squared norm.
By the definition of the operator norm:
\[
\|\Delta\|_2^2 = \left\| V W_m^\ell (\tilde{s}^{(1)} - \tilde{s}^{(2)}) \right\|_2^2.
\]

\textbf{Step 3:} Apply the norm inequality for linear transformations.
Since \(V W_m^\ell\) is a linear map from \(\mathbb{R}^d \to \mathbb{R}^C\), and \(\tilde{s}^{(1)} - \tilde{s}^{(2)}\) lies in \(\mathbb{R}^d\), we apply the norm inequality for linear transformations:
\[
\| \Delta \|_2^2 \geq \sigma_{\min}^2 \cdot \left\| \tilde{s}^{(1)} - \tilde{s}^{(2)} \right\|_2^2,
\]
where \(\sigma_{\min}\) is the smallest singular value of \(V W_m^\ell\).

\textbf{Step 4:} Apply the spectral norm assumption.
By assumption, \(V W_m^\ell\) is full-rank and has minimal singular value at least \(\sqrt{\epsilon}\), so:
\[
\sigma_{\min}(V W_m^\ell) \geq \sqrt{\epsilon}.
\]

\textbf{Step 5:} Conclude the bound.
Therefore:
\[
\| \Delta \|_2^2 \geq \epsilon \cdot \left\| \tilde{s}^{(1)} - \tilde{s}^{(2)} \right\|_2^2.
\]
This completes the proof.
\end{proof}
